MSstatsPTM: Statistical relative quantification of post-translational modifications in bottom-up mass spectrometry-based proteomics

### 2022-06-16 main/MSstatsPTM_Main.tex
Updated t-test and limma methods

\noindent {\bf Two-sample $t$-test} 
\todo{OV: could you help me with this part a little bit? I tried to organize the text, but could not quite finish. We should have the following order: (1) input, (2) model for the input, (3) model-based inference, (4) implementation; (5) good things and limitations. I think it may help to separate two versions: based on the features, and based on the ratios. Otherwise it is not very clear}
Many investigations perform differential abundance analysis of PTMs using two-sample $t$-test or its extensions. $t$-test takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features \todo{OV: let's separate the intensities and the ratios}, and compares the mean abundance of a PTM site from one condition to another~\cite{Schwammle2015}\cite{THOMAS2020}. The abundance in each run is taken as input and is often estimated by sum of peak intensities. For example, the log-abundance estimate for the PTM in Run $j$ of Condition $i$ is given by Equation \ref{eq:feature_estimation}. as input the adjusted PTM abundance of each condition and biological replicate, and estimates $\Delta$ in Equation \ref{eq:null_hyp}.
\todo{OV: what is the model?}
To estimate the parameters of interest in this trivial model, the approach takes as input

\begin{equation}
\begin{aligned}
\log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right).
\end{aligned}
\label{eq:feature_estimation}
\end{equation}

For adjustment with respect to unmodified peptides, the estimate of PTM abundance is divided by the protein abundance estimate, which is equivalent to take the difference of their log abundances. For condition $i$ and biological replicate $j$, the estimated quantity is denoted by $\hat{u}^{\ast}_{ij}$ and is given by Equation \ref{eq:adj_estimation}.

@@

to estimate the quantity of interest $\hat{\Delta}$ as well as its standard error, $\widehat{SE(\hat{\Delta})}$.  The statistical significance is determined by comparing the test statistic against the $t$ distribution, with degrees of freedom $df=2J-2$ in balanced designs.

While simple, these approaches do not fully account for all sources of variations, and are not directly applicable to experiments with complex designs. These include experiments with comparisons of multiple conditions, acquisition in multiple batches, and unbalanced designs. Additionally, while these approaches can be applied to experiments studying PTMs, there is not a self contained, straightforward implementation of the methods, making application challenging.

\medskip \noindent {\bf Limma} Modifications of $t$-test such as moderated $t$-test with Limma were also proposed~\cite{Ritchie_15a}\cite{Zhu}.

@@

To estimate $\hat{\Delta}$ in Figure \ref{fig:data-structure}, feature level summarization and unmodified protein adjustment are performed in the same way as $t$-test, shown in Equations \ref{eq:feature_estimation} and \ref{eq:adj_estimation}. Limma then fits a linear model using the same adjusted summarized features as in $t$-test, $\hat{u}^{\ast}_{11}$, $\hat{u}^{\ast}_{12}$, $\hat{u}^{\ast}_{21}$ and $\hat{u}^{\ast}_{22}$, to estimate $\hat{\Delta}$. To estimate $\widehat{SE(\hat{\Delta})}$, Limma leverages Empirical Bayes moderation to moderate the model variance of the individual PTM using the shared global variance of all PTMs modeled in the experiment. For a single PTM $m$ with an estimated fold change $\hat{\Delta}_m$, the estimated variance is calculated as $s^{adj}_{m} = P(s_{m}| s_\Theta)$ where $s_{m}$ is the variance resulting from the linear model of PTM $m$ and $s_\Theta$ is the global variance across all PTMs. The statistical significance is determined using $\hat{\Delta}_m$ and $\widehat{SE(\hat{\Delta})}$ calculated using the adjusted variance $s^{adj}_{m}$.

>>>>

\noindent {\bf Two-sample $t$-test} 
\todo{DK: I mainly explained the ratio version. I briefly mentioned the feature version, which is used to model ptms w/o adjustment. We could reference the versions back to hypotheses in Equation 1 and 2 but I didnt yet.}

Many investigations perform differential abundance analysis of PTMs using two-sample $t$-test or its extensions. $t$-test can be performed without accounting for confounding with global protein abundance by taking as input intensities of individual features from modified peptides, denoted by $y_{ijk}^{\ast}$. Additionally, it can be extended to remove confounding by taking as input intensity ratios of modified and unmodified peptide features and comparing the mean abundance of a PTM site from one condition to another~\cite{Schwammle2015}\cite{THOMAS2020}\cite{Mertins:2013}. This is done by taking the sum of peak intensities in each run for modified and unmodified features separately and dividing the estimate of PTM abundance by the protein abundance estimate. This is equivalent to taking the difference of their log abundances. For condition $i$ and biological replicate $j$, the estimated quantity is denoted by $\hat{u}^{\ast}_{ij}$ and is given by Equation \ref{eq:adj_estimation}.

@@

These estimated ratios are then used as input to the two sample $t$-test, the output of which is the estimate of our quantity of interest $\hat{\Delta}$ as well as the standard error, $\widehat{SE(\hat{\Delta})}$. The statistical significance is determined by comparing the test statistic against the $t$ distribution, with degrees of freedom $df=2J-2$ in balanced designs. $t$-test assumes that the data follows a normal distribution, sampled from a representative portion of the total population, and that the conditions of interest have equal variance.

While simple, these approaches do not fully account for all sources of variations, and are not directly applicable to experiments with complex designs. These include experiments with comparisons of multiple conditions, acquisition in multiple batches, and unbalanced designs. Additionally, because the adjustment is done on the replicate, or run, level, there must be a matching number of replicates in both the modified and unmodified experimental designs. If the designs are different, this method is no longer directly applicable. Finally, while these approaches can be applied to experiments studying PTMs, there is not a self contained, straightforward implementation of the methods, making application challenging.

\medskip \noindent {\bf Limma} Modifications of $t$-test such as moderated $t$-test with Limma were also proposed~\cite{Ritchie_15a}\cite{Zhu}\cite{Chappell:2021}.

@@

The inputs to Limma are the same as $t$-test. The intensities of individual features from modified peptides, $y_{ijk}^{\ast}$, can be used directly (in the presence of confounding with the unmodified protein). To remove confounding we transform the input by taking the intensity ratios of modified and unmodified peptide features, $\hat{u}^{\ast}_{ij}$, calculated in the same way as Equation \ref{eq:adj_estimation}. Limma then fits a linear model using the transformed inputs to estimate $\hat{\Delta}$. To estimate $\widehat{SE(\hat{\Delta})}$, Limma leverages empirical bayes moderation~\cite{Robbins1992} to moderate the model variance of the individual PTM using the shared global variance of all PTMs modeled in the experiment. For a single PTM $m$ with an estimated fold change $\hat{\Delta}_m$, the estimated variance is calculated as $s^{adj}_{m} = P(s_{m}| s_\Theta)$ where $s_{m}$ is the variance resulting from the linear model of PTM $m$ and $s_\Theta$ is the global variance across all PTMs. The statistical significance is determined using $\hat{\Delta}_m$ and $\widehat{SE(\hat{\Delta})}$ calculated using the adjusted variance $s^{adj}_{m}$.

\todo{DK: @Olga can you expand on the good things/limitations? I think I am missing some points here but not sure what exactly}

Limma provides an alternative that overcomes some of the limitations of $t$-test. First, utilizing empirical bayes moderation can increase the sensitivity of the analysis and is especially useful in cases of low replicates. Additionally, using a linear model allows for the sharing of variance information across comparisons of interest. This means we can more accurately measure the variance as the number of comparisons increases.

While a general improvement over $t$-test, there are still a number of limitations when using Limma. First, the method is limited in the types of experimental designs it can be applied to. The method is only directly applicable to experiments with at most two variance components, meaning that in more complex experiments it cannot account for all sources of variation. Additionally, it has the same issue as $t$-test with requiring the same experimental design for the modified and unmodified features, so that there are matching replicates in the ratio calculation. Finally, there is again no self contained implementation of the methods to PTMs, requiring manual transformation and application by the user at every step.
